---
title: "Mid-term project "
author: "Sultan Nawaz & MUHAMMAD USMAN KHALID"
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: true
    theme: cerulean
---

# Mid-term project GitHub repository URL:

<https://github.com/SultanNawaz/2024dsa_midproject_group7.git>\

# Introduction

With the increasing population, the world needs more food to feed on and for such purposes, intensive agriculture is taking place around the globe to feed and gain profits. To get more agricultural outputs many practices have been implemented and adopted by the agricultural and farming community. Using inputs like phosphatic fertilizers and chemicals including weedicides and pesticides to prevent weeds and pests are harmful to both soil and plants.

Phosphorus is an important nutrient to both plants and soil. Phosphorus has its pros and cons, therefore, its wide use is monitored to reduce environmental effects. The use of phosphorus in excess amounts is more often associated with Eutrophication, which is the leading threat to marine life.

With the increased use of phosphatic fertilizers, worldwide phosphorus reserves are declining at a very high speed. Researchers all over the globe are active in research to come up with a protocol where we could identify and use microbes to make P available which is usually bound to soil aggregates due to its complexes with Al and Fe in low pH soils and bound to Ca and Mg in alkaline soils. Much research published has witnessed the P-mineralization with the help of microbes including some bacteria and fungi. Keeping the same in mind we came up with a research where we will use some microbes and will use it with certain rates to assess what rates can be significant to use.

**Research gap:**

Previous research conducted on LEM does not give a suggestion/recommendation for the application rates to crops. Additionally, phosphorus behaves differently in different soil and environmental conditions. Applying LEM at different rates will enable us to understand the potential of microbes in nutrient uptake by plants which will add significance in reducing P levels of the soil. The data on P content in the soil after LEM application will be collected for the next 2 years, and will help us to assess whether or not LEM has helped us to lower the P concentration in the soil.

# Hypothesis and objectives

<<<<<<< HEAD
**Hypothesis**: Locally effective microbes will lower the phosphorus levels.

**Objectives:** The objective of our study is to determine phosphorus content with the application of locally effective microbes.
=======
Hypothesis: Locally effective microbes will lower the phosphorus levels.

Objectives: The objective of our study is to determine phosphorus content with the application of locally effective microbes.
>>>>>>> 3dcefc35e6b4895ccaffc47482881203e409a068

# Material and Methods

Site and liquid effective microbes preparation: This study was conducted at Horticulture farm in Watkinisville, Georgia in 2023 to test the effectiveness of locally effective microbes (LEM) in phosphorus draw-down at organic fields that was fertilized with organic inputs for 10 years. To make solid LEM, inuculum was collected from a hardwood, pine and mixed forest located in Watkinisville and was kept at in air tight container at room temperature for at least 4 weeks for fermentation after mixing it with crushed charcoal, yeast, 50 pounds of semolina, a gallon of molasses,unpasteurized goat milk and water. After fermentation, the liquid LEM was prepared by taking 3kg of solid LEM and hanging it in a 12 gallon bucket with a gallon of molasses and 8 gallons of water and keeping at room temperature for at least 3 weeks. This liquid LEM will be applied to the field.

Planting, LEM application and soil sampling:

Corn (wapsie) was planted at experimental site on May 15, 2022. LEM was applied at 0, 1.8 and 3.6 liter per meter square to the soil around the corn stems, whereas the soil samples were collected after 4 weeks of LEM application from a depth of 0-15cm with a soil probe.

## Study design

Randomized complete block design with repeated measure will be used in this project. Factor include locally effective microbes rates with three levels (0, 1.8L/m2 and 3.6L/m2) applied to 4 blocks making 36 experimental plots in total. Every block received each level of LEM 3 times as replicated thrice.

## Statistical analysis

For statistical analysis, RCB design was used with the help of R-software. The response variable in this study was phosphorus content in the soil whereas LEM rates were explanatory variables. The explanatory variables were treated randomly. The alpha level (P-value) was 0.05. Functions used during analysis were janitor, read.csv, summary, glimpse, mutate, geom_boxplot, lme4, lmer, geom_hline, aes, geom_jitter, geom_theme, options, augment, raneff, used from packages tidyverse, car, broom, ggplot, multcomp.

Linear assumptions including independence, normality, homeocedasity and outliers were checked and they did met.

## Running Code

##Here we will load the packages that we think will be used in this data set

```{r Packages}
#install.packages("tidyverse")
#install.packages("car")
#install.packages("lme4")
#install.packages("broom")
#install.packages("emmeans")
#install.packages("multcomp")

#Loading packaes to library

library(tidyverse) # for data wrangling and plotting
library(car) # for Anova function 
library(lme4) 
library(broom) # for model residuals extraction 
library(emmeans) # for model mean extraction
library(multcomp) # for pairwise comparison letter display

```
{##Here we are importing our data from the source file}
```{r data import}
      mid_df <- read.csv("../Data/Mid-Term Project.csv")

mid_df

```



##data imported looks okay to run for exploratory data analysis

# b) EDA tables
{##We are running code for explanatory data analysis to understand data,idetify relationship, check data quality etc.}
```{r summary_df}
summary(mid_df)
```

##The summary above shows that there are 9 treatments with LEM rates of 0, 1.8, and 3.6 where PO4 ranged from a minimum of 12.02 to a maximum of 147.81

{##WE will use glimpse here to see how our data looks like}
```{r glimpse}
glimpse(mid_df)
```

# c) Wrangling

{## Here come the process of gathering and transforming the data according to our required analysis.}
```{r mid_dfw}
mid_dfw <- mid_df %>%
  mutate(Replication = factor(Replication),
         LEM.rate..L.m2. = factor(LEM.rate..L.m2.)
         
         ) %>%
  janitor::clean_names()

mid_dfw
```

## Additionally we did use the Janitor function to clean the names and make it more presntable.  
```{r summary_dfw}

summary(mid_dfw)
mid_dfw
```


# d) EDA plots
{## In this chunk we presnted our data in plot (boxplot) where data points of data sets are presented with three different colors, each representing level/rate of LEM.}

```{r LEM boxplot}
ggplot(mid_dfw, aes(x = lem_rate_l_m2, 
                    y = po4_mg_kg,
                    color = lem_rate_l_m2)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")
```


##Box plots here suggest that LEM rate of Zero has higher values but we can confirm it after running the models and see ANOVA

# e) Statistical model

## ##Mixed-effect modeling in R

We will use a mixed effect model here to address heterogeneous variance, separate variance due to random sampling, accounting for random effects.

To account for more than one variance component (i.e., random effects and error), we can use function `lmer()` from package `lme4`.

```{r model}
# Changing to sum-to-zero contrast
options(contrasts = c("contr.sum", "contr.poly"))

# Model fitting
mid_mix_mod <- lmer(po4_mg_kg ~ lem_rate_l_m2 +
                       (1|replication),
                     data = mid_dfw
                       )

mid_dfw

# Summary
summary(mid_mix_mod)
```

# f) ANOVA table

```{## Here we are running ANOVA to identify significant effect of independent variables over dependent variables, test our hypothesis. ANOVA allow us determine significance based on p-value, and provide a framework for multiple comparisons.}


```{r ANOVA}
Anova(mid_mix_mod, type=3)
```

\## Anova table illustrated that there was no significant effect of our independent variables

A model is only valid for inference (i.e., means and pwc) IF it fulfills the linear model assumptions.

# g) Linear model assumptions

## Extracting residuals

\## Here we will extract our model residuals, and create studentized residuals.

The linear model assumptions are:

-   Independence (no pattern)\
-   Variance homogeneity (homoscedasticity)\
-   Normality\
-   Outlier detection (\< -3 or \> 3)

```{r mid_mix_resid}
library(broom.mixed)

mid_mix_resid <- augment(mid_mix_mod) %>%
  mutate(.studresid=rstudent(mid_mix_mod))

mid_mix_resid
```

One difference in mixed models is that the 3 first assumptions are also applied to the **random effects**, so we need to check it for them as well.

### Random effects are iid \~ N(0,var_a)

```{r randeff_rep}
randeff_rep <- ranef(mid_mix_mod)$rep

randeff_rep
```

For random effects with so few levels (i.e., 4 since that's the number of blocks), the QQ plot is one of the only ways to check for assumptions on random effects.

```{r QQ plot for random effects}
ggplot(randeff_rep, aes(sample=`(Intercept)`))+
  stat_qq(  shape = 20,
            fill = "blue", 
            size = 3,
            alpha = .8
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```

Nothing to worry here, especially since we only have 4 points.

### Within-group errors are iid \~ N(0, var_e)

```{r }
ggplot(mid_mix_resid, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="red")+
  geom_point(shape = 20,
             fill = "blue", 
             size = 3,
             alpha = .7) +
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "red")+
  theme_bw()
```

-   **Independence**: seems liker there is a potential pattern, but since error bands comprise 0 on the y-axis, I'm not worried.

-   **Variance homoscedastic**: seems constant across the x-axis.

-   **Outliers**: nothing outside the -3,3 boundaries.

```{r}
ggplot(mid_mix_resid, aes(sample=.studresid))+
  stat_qq(  shape = 20,
            fill = "blue", 
            size = 3,
            alpha = .8
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```

```{r}
ggplot(mid_mix_resid, aes(x=.studresid))+
  geom_density(color = "black",
               fill = "blue",
               alpha = .8)+
  scale_x_continuous(breaks = c(-3,0,3), limits = c(-3,3))+
  theme_bw()

```

-   **Normality**: residuals seem normal.

# g) Final plot

##After going through the linear assumption we will run codes to make the final plot to inspect data distribution of both raw data and statistical model summary with letter separation. \

```{r mid final plot}
ggplot(mapping = aes(fill = lem_rate_l_m2))+
  # Raw data and boxplots  
  geom_boxplot(data = mid_dfw,
               aes(x = lem_rate_l_m2, y = po4_mg_kg),
               alpha = .8) +
  geom_jitter(data = mid_dfw,
               aes(x = lem_rate_l_m2, y = po4_mg_kg),
              shape = 20,
              size = 3,
              alpha = .7) +
   
  labs(x = "LEM rate (l/m2)",
       y = "PO4 (mg/kg)") +
  scale_fill_viridis_d() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none")

ggsave("../Output//mid_mix_means.png",
       width = 8,
       height = 5)
```

##Final plot illustrates no significance at all hence box plots are not labelled with the letters. Concluding that LEM as an independent variable has not shown any significant effect in lowering the P content of soil during the study period.
